Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-29 manuscript/manuscript.tex
more SV edits

\begin{document}
\maketitle
\begin{abstract}
Recent work has produced conflicting interpretations on the question of
whether adults undergo durable changes in their personal culture over
time. This paper asserts that this divergence is less due to
researchers' findings than the approach taken -- a tournament of models
focusing on the presence or absence of change. To move beyond this, we
reanalyze the data used previously in this debate with an approach that
quantifies the processes of cultural difference. In doing so, we compare
how much cultural difference is explained by intrapersonal change and
interpersonal differences. Our results harmonize recent findings by
showing that, while most measures of personal culture show durable
change over time, the variance explained by intrapersonal change is
typically small compared to that of interpersonal differences at
baseline. Quantifying these processes offers new ways to address
theoretical questions of when and how their relative importance shifts.
We exemplify this approach by exploring how the importance of these
processes differs between college graduates and the rest of the
population.
\end{abstract}

@@

Does personal culture---an individual's attitudes, beliefs, values, and
practices---change over the life course, or is it largely fixed by
adulthood? This question underlies an important contemporary debate in
sociology (Kiley and Vaisey 2020; Lersch 2023; Lizardo 2017) and has
deep roots in seemingly contradictory theoretical perspectives. For
example, pragmatist theories of action claim that changes in social
environments cause people to adapt their views and make new cultural
meanings (Gross 2009; Swidler 2001), while Bourdieusian practice
theories argue that the ``past conditions of production'' leave a mark
on people's personal culture that lasts throughout their lives (Bourdieu
1990). Models of social influence assume that people adapt their culture
in the face of new information (Goldberg and Stein 2018), while the
emphasis on cohort effects in models of aggregate social and cultural
change requires them to be open to change while young but become fairly
resistant to it as they age (Ryder 1965). Finally, life course theories
posit important changes over time as people advance through important
transitions in their lives (Bardi et al. 2009; Elder, Johnson, and
Crosnoe 2003).

Because all of these theoretical perspectives have some empirical
support, the current debate is not about whether their posited processes
exist at all, but about their relative contribution to explaining the
cultural differences we see in the world. That said, scholars have
struggled to reach a consensus on the importance of change during
adulthood. Over the span of a few years, most apparent changes in
people's responses to personal survey items appear to be transitory,
with little evidence of persistent change among adults (Kiley and Vaisey
2020; Vaisey and Kiley 2021). This suggests that change during adulthood
is not a major factor in explaining personal cultural differences. When
we consider a longer time horizon, however, there is evidence that
adults make at least some persistent changes (Lersch 2023).

These seemingly inconsistent findings are partly due to the fact that
researchers have taken a ``tournament of models'' approach to adjudicate
different theoretical perspectives (Lersch 2023: p.~228). That is,
researchers have compared the fit of different statistical models to
each item to determine whether it is more consistent with ``no change''
or ``any change.'' The size of these piles then serves as the primary
evidence for the truth of the theory (Kiley and Vaisey 2020; Lersch
2023; Vaisey and Kiley 2021).

We argue, however, that asking \emph{whether} people change on an item
may cause more problems than it solves. Being able to detect \emph{any}
change on an item depends not only on sample populations, measurement,
statistical methods, and power, but also on definitions of what counts
as ``change.'' These differences can limit the possibility of agreement
on which ``pile'' a variable belongs. Moreover, asking whether there is
evidence of change reduces to, ``Is there \emph{any} evidence of change
in this item?'' But such binary inquiries are ill-suited to address the
core of the theoretical debate: determining \emph{how much} the process
of intrapersonal change contributes to cultural differences.

There is enough evidence to assert that neither pre-adult socialization
nor the contemporary social context alone can explain the full range of
cultural differences. The question that remains is about the
\emph{relative} contributions of these processes. In the current debate,
for a single construct (e.g., support for gay rights), existing
approaches can only make claims about \emph{whether} intrapersonal
change is happening, not about how much. This also prevents
investigating under what conditions, across which groups, and in what
domains do we see differences in the relative importance of
intrapersonal change and pre-existing interpersonal differences. These
questions are about \emph{degree}, not existence; to take this debate
forward therefore requires precise measurement rather than declarations
of victory for a particular perspective.

In this paper, we move beyond the ``tournament of models'' approach and
introduce a method for quantifying the relative contributions of
interpersonal differences and intrapersonal change for a single item.
Drawing on seven panel data sets from five countries, we quantify the
amount of variance in survey responses that is due to between-person
differences at a single point during the survey versus the amount of
variance in responses that is due to systematic intrapersonal change
over the study period. This measure quantifies the amount of change over
the duration of the panel relative to the stable differences that exist
between people.

Using this measure, we observe similar patterns across the datasets that
have previously yielded conflicting interpretations. In all
datasets---despite their varied social contexts and duration---we find a
similar distribution of intrapersonal change. In general, we find that
intrapersonal change accounts for a substantially smaller amount of
variance in personal culture survey item than stable interpersonal
differences. Of course, whether the amount of intrapersonal change in an
item is large enough to be ``meaningful'' is a substantive and
theoretical question. Nevertheless, on many measures of personal culture
(some observed for over a decade), we do not observe enough
intrapersonal change for it to play a substantial role in explaining the
differences we observe between adults. Notable, interesting, and
important exceptions exist, of course, but they are departures from the
overall pattern.

In addition to advancing a theoretical debate, our method can also help
address substantive questions. To illustrate how it might be used by a
researcher with a specific topical interest, we investigate differences
in the relative importance of intrapersonal change for people with and
without a college degree. We find a smaller amount of systematic change
among college-educated respondents, suggesting that college solidifies
one's personal culture rather than fostering an openness to new
information. This offers a novel empirical basis to theorize about the
role of university education in shaping personal culture.

Taken together, the goal of this paper is to advance---and, hopefully,
transcend---debates about \emph{whether} adults change. People do
change, at least a little, on most things. Our method can help quantify
exactly \emph{how much}.

@@

\hypertarget{change-and-stability-in-personal-culture}{%
\subsection{Change and Stability in Personal
Culture}\label{change-and-stability-in-personal-culture}}

@@

There is no reason to believe that only one of these two models is
``correct'' at all times and in all places. A population observed over
some period of time contains a mix of people who are noticeably changing
and people who are not. Instead, different perspectives argue that each
of these ideal-typical models is more operative at different times, for
different people, and for different elements of personal culture. For
example, adolescence and early adulthood is typically viewed as a
``formative period'' for personal cultural development and thus
characterized by higher rates of active updating, while middle age and
later life are potentially characterized more by settled dispositions
(Alwin and Krosnick 1991; Eaton et al. 2009; Krosnick and Alwin 1989).
Similarly, salient issues, such as views around gay rights in the 2010s;
issues that see substantial elite realignments, such as views around the
Vietnam conflict in the 1970s (Zaller 1992); and novel issues of public
opinion, such as views around vaccines during the Covid-19 pandemic
(Scoville et al. 2022), might be characterized by active updating, while
established issues of low salience are characterized by stability.

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the
majority of items preferred the AUM, the amount of durable change
detected on these items was small. A substantial minority of questions
(39 percent) favored the SDM, meaning they showed no evidence of durable
change. Questions with more evidence of durable change included salient
issues like gay marriage and questions tapping ``public'' statements or
behaviors such as partisan identification and religious service
attendance. There was also more evidence of durable change among early
adults (people ages 18-30) than the rest of the population. Overall, the
researchers concluded that ``results ultimately suggest that real,
persistent attitude change is an uncommon phenomenon among adults'\,'
(Kiley and Vaisey 2020: p.~500; see also Vaisey and Kiley 2021). This
lack of durable change is consistent with other recent findings that
cohort replacement plays a somewhat larger role than period effects in
explaining differences in personal culture (Vaisey and Lizardo 2016).

@@

Despite differences in how these researchers interpret their findings,
the empirical results are not far apart. Kiley and Vaisey (2020) found
that the majority of items they tested favored the AUM, meaning there
was evidence of durable change among adults, a pattern consistent with
Lersch's results. Also Lersch (2023) found evidence of durable change on
most items, but this change only accounted for a small proportion of
variance explained. On average, people changed only about .07 standard
deviations over 10 years. This is consistent with Kiley and Vaisey's
finding that durable change is often small. Moreover, a quarter of the
items Lersch studied still preferred the SDM, suggesting that even with
different assumptions assumptions about change and more extensive data,
many questions show no evidence of durable change. Consistent with this,
in studies showing evidence of durable change in specific measures of
personal culture, the change is often small. For example, Brocic and
Miles (2021) estimate that completing graduate degrees in humanities,
arts, and social sciences shift peoples' moral relativism only about .2
standard deviations on average compared to people with no college
degree, and this is the largest effect they identify. And studies of
aggregate change show that, even on items where cohort effects explain
more variance than period effects, there is always evidence that some
people change over time (Vaisey and Lizardo 2016). In other words,
despite different interpretations, the results of previous work are in
many ways highly consonant.

@@

The current debate can be simplified to asking whether people
\emph{ever} change. This is rarely theoretically interesting since the
answer is almost always ``yes.'' Over sufficient time and in a large
enough sample, researchers will likely observe some evidence for durable
individual change. A lack of evidence for change may be due to a poor
survey question, a low resolution in response options, or that the
question simply was not asked for long enough or to enough people.
Conversely, finding evidence for some change tells us only that it was
``not zero.'' From that alone, we learn little about how much change has
happened in a population.

A more theoretically productive approach begins with a model assuming
that during an observed time period people might remain stable, might
change a little, or might undergo significant shifts in their personal
culture. Lersch's LCAM does this by modeling each individual as
following a personal linear trajectory. But rather than debating whether
this model fits a particular data set better than a model that assumes
that people never change, researchers should quantify these individual
changes and compare those to the differences that exist between people
at the panel's start. In other words, the focus should shift from
asking, ``do people change?'' to asking, ``what are the relative
contributions of change and initial differences for explaining variance
in personal culture?''

We propose an approach that offers such a quantification. Our measure
relates two metrics that each reflect a combination of relevant
theoretical processes. The first metric, interpersonal differences at
the start of the survey, reflects the accumulated experiences of people
prior to entering the panel survey that shape how they answer questions.
Lersch calls this ``early imprinting.'' While commonly associated with
experiences during a formative period that result in settled
dispositions, this variation could also reflect experiences that
happened at any stage in life as long as they predate the panel and
consistently affect subsequent responses. For instance, for those people
entering the panel post-retirement, this ``imprinting'' might reflect
this pivotal life transition. Consequently, this metric also reflects
variation in individuals' social roles or statuses at the start of the
panel that were important in shaping their dispositions.

The second metric, the amount of intrapersonal linear change people make
during the panel, captures durable changes in personal culture over
time. This captures the set of processes collectively called
``persistent change'' or ``adaption'' by Lersch and ``active updating''
by Kiley and Vaisey. Lersch attributes these changes to social triggers
such as moving into a new environment or adopting new social roles
(although he never measures these directly). Additionally, they might
reflect the diffusion of new cultural forms across social networks, cues
from political elites or otherwise culturally influential leaders, the
emergence of issues in politics or culture, or large-scale social
shifts.

A third metric, which we might call ``residual variance'' or
``fluctuation,'' accounts for the remainder of variance in peoples'
responses. These non-durable changes emerge for a variety of reasons.
For example, people might not have a clear disposition on a particular
item as it is asked. Instead, they might internalize a broad set of
considerations and construct an opinion in the context of the survey
interview, with different considerations coming to the forefront of
their cognition during each interview (Feldman and Zaller 1992;
Tourangeau, Rips, and Rasinski 2000; Zaller 1992). This variance can
also include measurement error, such as misinterpreted questions,
erroneous response selections, or responses getting coded incorrectly.
Although this third metric at least partially reflects important
processes of personal culture, it does not directly touch on the ongoing
debate outlined here. Because of this, we quantify this component but
focus principally on the other two.

At a theoretical level, the coexistence of the sets of processes
captured in the two main metrics, interpersonal differences and
intrapersonal change, is undeniable, and they are linked in many ways.
Any intrapersonal change during one's life will likely manifest as
interpersonal differences by the time people enter a panel survey.
Furthermore, unless people are entirely socialized early on and never
deviate from these dispositions, we expect observing intrapersonal
change in a segment of the population when surveyed over time.

Furthermore, these metrics are not necessarily useful in isolation or
without context. What is important for the theoretical debate is the
relative contribution of these two components in explaining cultural
differences in a population. In other words, when we look at a
population of adults over some time frame, is there enough intrapersonal
change relative to interpersonal difference to view it as an important
explanation for why people differ from each other as observed? Only
through quantifying the two processes we can get closer to a true answer
to this question.

Yet the true utility of the proposed quantification goes beyond settling
past debates; it allows researchers to ask questions that were difficult
to explore with previous approaches. For example, classifying survey
items based on whether they show durable change overlooks possibly
important differences among those that do. By quantifying the relative
contributions of interpersonal differences and intrapersonal change,
researchers can gauge their relative importance in explaining
differences among survey items, between groups within a given
population, across time, and across societies.

@@

By shifting the question and adjusting the approach, we expect
consistent patterns across datasets previously analysed in the debate.
First, because previous work found evidence of durable change on most
measures of personal culture, we expect that we will find some
intrapersonal change in most items. But because these earlier studies
also found many questions (between 25 and 35 percent) lacking evidence
of \emph{any} durable change, we expect considerable variation in the
variance accounted for by intrapersonal change, with some questions
indicating essentially no variance attributable to such change.

Second, we expect the variance attributable to intrapersonal change to
be modest compared to interpersonal differences at the start of the
survey. This expectation should be uncontroversial, regardless of the
theoretical process assumed to be most relevant for a particular
question. Panel surveys do not encompass individuals' entire lifetimes.
Thus, even if intrapersonal change during adulthood is the major driver
of cultural differences, the change detected in a panel covering 20
years probably will not fully account for the entire variance of
interpersonal differences.

Our expectations so far largely echo what was observed in previous
studies. However, theoretical perspectives differ in how much variance
they would expect to be driven by intrapersonal change, enabling
quantification to provide a new perspective on existing questions.
``Formative period'' and cohort theories, which emphasize early-life
socialization as a strong influence on adult personal culture, suggest
that the variance from intrapersonal change would still fall short of
explaining interpersonal differences, even when we account for the
duration of observation. Conversely, other theoretical perspectives
emphasizing the importance of people's contemporaneous social
environment for personal culture might expect intrapersonal change to
account for a large portion of the variance, regardless of the duration.

Third, we expect the variance explained by each process ---interpersonal
differences and intrapersonal change--- to depend on specific attributes
of the question and panel. First, the higher the response resolution for
a given question and the larger the sample, the greater the variance
explained by intrapersonal change should be, as smaller changes in
intrapersonal change will be easier to detect. Second, the LCAM and
various theoretical perspectives that underlie it posit that social
transitions drive changes in personal culture. Since the probability
that any person experiences such a transition goes up the longer we
observe them, we expect panel duration to be positively related to the
proportion of variation explained by intrapersonal change.

@@

We use data from seven nationally representative panel surveys from
Australia, Germany, Great Britain, Switzerland, and the United States
(summarized in Table 1), combining all the data files used in previous
work (Kiley and Vaisey 2020; Lersch 2023).\footnote{For more information
  on these data sources, see (Goebel et al. 2019; Income Dynamics 2013;
  Smith et al. 2022; Summerfield et al. 2011; Taylor 1996; University of
  Essex and Research 2019; Voorpostel et al. 2016).} These studies cover
a long period of time (the range of surveys spans from 1968 to 2021),
with 609 personal culture items that capture attitudes, beliefs, values,
self-assessments, self-descriptions, and behaviors (Alwin 2007). We
restricted the sample such that individuals between the ages of 18 and
79 are included without further elimination, and in all surveys, we used
all possible cases for which respondents provided responses. In the end,
the analyses that follow rely on a cross-national sample with a
cross-domain set of items to capture a broad range of individual
personal culture. Supplemental Materials A documents the list of all
variables used in the upcoming analyses.

@@

We start with Lersch's (2023) Life Course Adaption Model, which
formalizes survey responses at time \(t\) as a function of
individual-level random intercepts and slopes for survey age, what is
often called a mixed-effects growth curve model. This model formally
assumes a set of propositions about change that reflect the theoretical
debate to this point. First, consistent with the settled disposition
model, it assumes that people start the survey with cultural
differences, modeled as random intercepts for each respondent. Second,
it assumes that people change over time, taking the form of random
slopes for each respondent as a linear function of time. Third, it
assumes that people deviate around this baseline randomly over time,
reflecting ``fluctuation'' or short-term non-persistent change.
Formally, this can be written as

@@

As a first step in our analysis, we fit this LCAM to each of the 609
measures of personal culture outlined above. We then use these models to
derive two measures of systematic variance that correspond to the two
collections of theoretical processes outlined above: interpersonal
differences at baseline and intrapersonal change over time.

We estimate these using a form of variance decomposition. First, to
measure stable interpersonal differences, we calculate the sum of the
squared residuals derived from assuming that each person spends the
entire survey at the estimated midpoint of their LCAM
trajectory.\footnote{We use the midpoint estimate, rather than the value
  at the first wave, because this is the best measure of ``baseline''
  available under the assumption that the person does not change. The
  first wave measure alone contains an unknown amount of measurement
  error and transient fluctuations.} We then divide this value by the
total sum of squares and subtract this number from from 1, producing a
value we denote as \(V(D)\). This is the proportion of total variance
attributable to interpersonal differences. (This number is equivalent to
the conditional \(R^2\) value generated by a model with only random
intercepts for all respondents.) Second, we calculate another sum of
squared residuals generated by allowing the predictions to vary by
estimated LCAM trajectory. We divide this number by the total sum of
squares and subtract it from 1. We then subtract \(V(D)\) from this
value. This gives us a value we denote as \(V(C)\), the additional
proportion of variance accounted for by intrapersonal change.

@@

The most notable result in the meta analysis is the negative coefficient
attached to the duration of years covered by the question. The longer a
question is observed, the less is the value of \(\omega\). Theories of
personal cultural change that link changes in personal culture to social
experiences, including the LCAM, suggest that the longer we observe
respondents, the more likely people are to undergo potentially
transformative experiences and therefore the more variance would be
attributable to intrapersonal change. Finding a negative coefficient
here seems to challenge to that assumption.

@@

>>>>

\begin{document}
\maketitle
\begin{abstract}
Does personal culture---a person's attitudes, beliefs, values, and
practices---change over the life course, or is it largely fixed by
adulthood? Recent work has reached substantially different conclusions
on this question. We argue that this disagreement is an unintended
consequence of the ``tournament of models'' approach researchers have
used. We introduce a new measure for quantifying the proportion of
systematic variance in panel data attributable to intrapersonal change
versus stable interpersonal differences. Applying this measure to 609
items from seven surveys in five countries, we find that although
intrapersonal change is common, it is generally small in magnitude. We
also demonstrate that this measure is useful for comparing social groups
by showing that intrapersonal change is less common among U.S. college
graduates than among those without a college degree. Overall, we believe
this new measure will be a useful methodological tool and help reconcile
an important theoretical debate.
\end{abstract}

@@

Does personal culture---a person's attitudes, beliefs, values, and
practices---change over the life course, or is it largely fixed by
adulthood? This question underlies an important contemporary debate in
sociology (Kiley and Vaisey 2020; Lersch 2023; Lizardo 2017) and has
deep roots in seemingly contradictory theoretical perspectives. For
example, pragmatist theories of action claim that changes in social
environments cause people to adapt their views and make new cultural
meanings (Gross 2009; Swidler 2001), while Bourdieusian practice
theories argue that the ``past conditions of production'' leave a mark
on people's personal culture that lasts throughout their lives (Bourdieu
1990). Models of social influence assume that people adapt their culture
in the face of new information (Goldberg and Stein 2018), while the
emphasis on cohort effects in models of aggregate social and cultural
change requires them to be open to change while young but become fairly
resistant to it as they age (Ryder 1965). Finally, life course theories
posit important changes over time as people advance through important
transitions in their lives (Bardi et al. 2009; Elder, Johnson, and
Crosnoe 2003).

Because all of these perspectives have some empirical support, the
current debate is not about whether the processes they posit exist at
all, but about their relative contribution to explaining the cultural
differences we see in the world. Nevertheless, researchers have
struggled to reach a consensus on the importance of cultural change
during adulthood. Over the span of a few years, most apparent changes in
adults' survey responses appear to be transitory, with little evidence
of persistent change (Kiley and Vaisey 2020; Vaisey and Kiley 2021).
This suggests that change during adulthood may not be a major factor in
explaining contemporary cultural divides. By contrast, when we consider
a longer time horizon, there is evidence that adults make at least some
persistent changes (Lersch 2023).

These seemingly inconsistent findings are partly due to the fact that
researchers have taken a ``tournament of models'' approach to adjudicate
different theoretical perspectives (Lersch 2023: p.~228). That is,
researchers have relied on model selection criteria---primarily the
Bayesian Information Criterion (Raftery 1995)---to sort variables into
the ``no change'' or ``any change'' piles. The size of these piles then
serves as the primary evidence for the truth of the theory (Kiley and
Vaisey 2020; Lersch 2023; Vaisey and Kiley 2021).

We argue, however, that asking \emph{whether} people change on an item
is an unintended source of confusion. Being able to detect \emph{any}
change on an item depends not only on sample populations, measurement,
statistical methods, and power, but also on definitions of what counts
as ``change.'' These differences can limit the possibility of agreement
on the assignment of a variable to the right pile. Moreover, asking
whether there is evidence of change reduces to, ``Is there \emph{any}
evidence of change in this item?'' But such binary questions are
ill-suited to address the core of the theoretical debate: determining
\emph{how much} the process of intrapersonal change contributes to
cultural differences.

Of course neither pre-adult socialization nor the contemporary social
context alone can explain the full range of cultural differences. The
relevant question is about the \emph{relative} contributions of these
processes. In the current debate, for a single construct (e.g., support
for gay rights), model comparison approaches can only help decide
\emph{whether} intrapersonal change is happening, not about how much.
This also prevents investigating under what conditions, across which
groups, and in what domains we might see differences in the the amount
of intrapersonal change. These questions are about \emph{degree}, not
existence. Taking this debate forward therefore requires precise
measurement rather than declarations of victory for a particular
perspective.

In this paper, we move beyond the ``tournament of models'' approach and
introduce a new method for quantifying the relative contributions of
interpersonal differences and intrapersonal change for a single item. We
use 609 items from seven panel surveys from five countries to quantify
the proportion of variance in that is attributable to systematic
intrapersonal change over the study period rather than to stable
interpersonal differences.

We observe similar results across all datasets despite their varied
social contexts and duration. In general, we find that intrapersonal
change accounts for a small amount of variance in personal culture
survey items. Of course, any question about whether a given amount of
intrapersonal change is ``meaningful'' or ``important'' is a substantive
and theoretical. Nevertheless, on many measures (some observed for over
a decade), we do not see enough change to believe that intrapersonal
change processes play a substantial role in explaining the differences
we observe between adults. Notable, interesting, and important
exceptions exist, but they are departures from the overall pattern.

In addition to advancing a theoretical debate, our method can also help
address specific substantive questions. To illustrate how it might be
used by a researcher with a specific topical interest, we investigate
differences in the relative importance of intrapersonal change for
people with and without a college degree. We find a smaller amount of
systematic change among college-educated respondents, suggesting that
college crystallizes one's personal culture rather than fostering an
openness to new information. This offers a new empirical basis to
theorize about the role of education in influencing personal culture.

Taken together, our goal in this paper is to advance---and, hopefully,
transcend---debates about \emph{whether} adults change. People do
change, at least a little, on most things. Our method can help quantify
exactly \emph{how much}.

@@

\hypertarget{stability-and-change-in-personal-culture}{%
\subsection{Stability and Change in Personal
Culture}\label{stability-and-change-in-personal-culture}}

@@

There is no reason to believe that only one of these two models is
``correct'' at all times and in all places. A population observed over
some period of time contains a mix of people who are changing and people
who are not. Instead, different perspectives argue that each of these
ideal-typical models is more operative at different times, for different
people, and for different elements of personal culture. For example,
adolescence and early adulthood is typically viewed as a ``formative
period'' for personal cultural development and thus characterized by
higher rates of active updating, while middle age and later life are
potentially characterized more by settled dispositions (Alwin and
Krosnick 1991; Eaton et al. 2009; Krosnick and Alwin 1989). Similarly,
salient issues, such as views around gay rights in the 2010s; issues
that see substantial elite realignments, such as views around the
Vietnam conflict in the 1970s (Zaller 1992); and novel issues of public
opinion, such as views around vaccines during the Covid-19 pandemic
(Scoville et al. 2022), might be characterized by active updating, while
established issues of low salience are characterized by stability.

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the
majority of items did prefer the AUM, the amount of durable change
detected on these items was small. A substantial minority of questions
(39 percent) favored the SDM, meaning they were more consistent with
zero durable change. Questions with more evidence of durable change
included salient issues like gay marriage and questions tapping
``public'' statements or behaviors such as partisan identification and
religious service attendance. There was also more evidence of durable
change among early adults (people ages 18-30) than among the rest of the
population. Overall, the researchers concluded that ``results ultimately
suggest that real, persistent attitude change is an uncommon phenomenon
among adults'\,' (Kiley and Vaisey 2020: p.~500; see also Vaisey and
Kiley 2021). This lack of durable change is consistent with other recent
findings that cohort replacement plays a somewhat larger role than
period effects in explaining differences in personal culture (Vaisey and
Lizardo 2016).

@@

Despite differences in how these researchers interpret their findings,
the empirical results are not far apart. Kiley and Vaisey (2020) found
that the majority of items they tested favored the AUM, meaning there
was evidence of durable change among adults, a pattern consistent with
Lersch's results. Also Lersch (2023) found evidence of durable change on
most items, but these changes were small. On average, people changed
only about .07 standard deviations over 10 years. This is consistent
with Kiley and Vaisey's finding that durable change is often small.
Moreover, a quarter of the items Lersch studied still preferred the SDM,
suggesting that even with different assumptions about change and more
extensive data, many questions still are consistent with a model that
allows zero intrapersonal change.

Other studies are compatible with these findings as well. For example,
Brocic and Miles (2021) estimate that completing graduate degrees in
humanities, arts, and social sciences shift peoples' moral relativism
only about .2 standard deviations on average compared to people with no
college degree, and this is the largest effect they identify. And
studies of aggregate change show that, even on items where cohort
effects explain more variance than period effects, there is always
evidence that some people change over time (Vaisey and Lizardo 2016). In
other words, despite different interpretations, the results of previous
work are in many ways highly consonant.

@@

To this point, the theoretical debate has been about whether people
\emph{ever} change. As we argued above, this is rarely the right
question since the answer is almost always ``yes.'' Over sufficient time
and in a large enough sample, researchers will likely observe some
evidence for durable individual change. A lack of evidence for change
may be due to a poor survey question, a low resolution in response
options, or that the question simply was not asked for long enough or to
enough people. Conversely, finding evidence for some change tells us
only that it was ``not zero.'' From that alone, we learn little about
how much intrapersonal change has happened in a population.

A more theoretically productive approach begins with a model assuming
that during an observed time period people might remain stable, might
change a little, or might undergo significant shifts in their personal
culture. Lersch's LCAM does this by modeling each individual as
following a personal linear trajectory. But rather than debating whether
this model fits a particular data set better than a model that assumes
that people never change, it would be better to quantify these
individual changes and compare them to the stable differences that exist
between people. In other words, the focus should shift from asking, ``do
people change?'' to asking, ``what are the relative contributions of
change and pre-existing differences for explaining variance in personal
culture?''

We propose an approach---which we formalize below---that offers such a
quantification. Our measure relates two variance components that each
reflect a combination of relevant theoretical processes. The first
component, stable interpersonal differences, reflects the accumulated
experiences of people prior to entering the panel survey. Lersch (2023:
24) calls this ``early imprinting.'' While commonly associated with
experiences during a formative period that result in settled
dispositions, this variation could also reflect experiences that
happened at any time as long as they predate the panel and consistently
affect subsequent responses. For instance, for those people entering the
panel post-retirement, this ``imprinting'' might reflect this pivotal
life transition. Consequently, this component also reflects variation in
individuals' social roles or statuses at the start of the panel that
were important in shaping their dispositions.

The second component, the amount of intrapersonal linear change people
make during the panel, captures durable changes in personal culture over
time. This captures the set of processes collectively called
``persistent change'' or ``adaption'' by Lersch and ``active updating''
by Kiley and Vaisey. Lersch attributes these changes to social triggers
such as moving into a new environment or adopting new social roles
(although he does not measure these directly). They might also reflect
the diffusion of new cultural forms across social networks, cues from
political elites or otherwise culturally influential leaders, the
emergence of issues in politics or culture, or large-scale social
shifts.

A third component, which we might call ``residual variance'' or
``fluctuation,'' accounts for the remainder of variance in peoples'
responses. These non-durable changes emerge for a variety of reasons.
For example, people might not have a clear disposition on a particular
item as it is asked. Instead, they might internalize a broad set of
considerations and construct an opinion in the context of the survey
interview, with different considerations coming to the forefront of
their cognition during each interview (Feldman and Zaller 1992;
Tourangeau, Rips, and Rasinski 2000; Zaller 1992). This variance can
also include measurement error, such as misinterpreted questions,
erroneous response selections, or responses getting coded incorrectly.
Although this third component at least partially reflects important
processes of personal culture, it does not directly touch on the ongoing
debate outlined here. Because of this, we quantify this component but
focus principally on the other two.

At a theoretical level, the coexistence of the sets of processes
captured in the two main components, interpersonal differences and
intrapersonal change, is undeniable, and they are linked in many ways.
Any intrapersonal change during one's life will likely manifest as
interpersonal differences by the time people enter a panel survey.
Furthermore, unless people are entirely socialized early on and never
deviate from these dispositions, we expect observing intrapersonal
change in a segment of the population when surveyed over time.

Furthermore, these components are not necessarily useful in isolation or
without context. What is important for the theoretical debate is the
relative contribution of these two components in explaining cultural
differences in a population. In other words, when we look at a
population of adults over some time frame, is there enough intrapersonal
change relative to interpersonal difference to view it as an important
process for why people differ from each other? Only through quantifying
the two components we can get closer to a true answer to this question.

Yet the true utility of the proposed quantification goes beyond settling
past debates; it allows researchers to ask new questions. For example,
classifying survey items based on whether they show durable change
overlooks possibly important differences among those that do show such
change. By quantifying the relative contributions of interpersonal
differences and intrapersonal change, researchers can gauge their
relative importance in explaining differences among survey items,
between groups within a given population, across time, and across
societies.

@@

By shifting the question and adjusting the approach, we expect
consistent patterns across datasets. First, because previous work found
evidence of durable change on most measures of personal culture, we
expect that we will find some intrapersonal change in most items. But
because these earlier studies also found many questions (between 25 and
35 percent) lacking evidence of \emph{any} durable change, we expect
considerable variation in the proportion of variance accounted for by
intrapersonal change, with some questions indicating essentially none at
all.

Second, we expect the variance attributable to intrapersonal change to
be modest compared to pre-existing interpersonal differences. This
expectation should be uncontroversial, regardless of the theoretical
process assumed to be most relevant for a particular question. Panel
surveys do not encompass individuals' entire lifetimes. Thus, even if
intrapersonal change during adulthood is the major driver of cultural
differences, the change detected in a panel covering 20 years probably
will not fully account for the entire variance of interpersonal
differences.

Our expectations so far largely echo what was observed in previous
studies. However, theoretical perspectives differ in how much change
they expect, enabling quantification to provide a new perspective on
existing questions. ``Formative period'' and cohort theories, which
emphasize early-life socialization as a strong influence on adult
personal culture, imply low amounts of intrapersonal change, even when
we account for the duration of observation. Conversely, other
theoretical perspectives emphasizing the importance of people's
contemporaneous social environment might expect a larger degree of
intrapersonal change, even in a relatively short panel.

Third, we expect the variance explained by each process---interpersonal
differences and intrapersonal change---to depend on specific attributes
of the question and panel. First, the higher the response resolution for
a given question and the larger the sample, the greater the variance
attributed to intrapersonal change should be, as smaller changes in
intrapersonal change will be easier to detect. Second, life course
theories posit that social transitions drive changes in personal
culture. Since the probability that any person experiences such a
transition goes up the longer we observe them, we expect panel duration
to be positively related to the proportion of variation explained by
intrapersonal change.

@@

We use data from seven nationally representative panel surveys from
Australia, Germany, Great Britain, Switzerland, and the United States
(summarized in Table 1), combining all the data files used in previous
work (Kiley and Vaisey 2020; Lersch 2023).\footnote{For more information
  on these data sources, see (Goebel et al. 2019; Income Dynamics 2013;
  Smith et al. 2022; Summerfield et al. 2011; Taylor 1996; University of
  Essex and Research 2019; Voorpostel et al. 2016).} These studies cover
a long period of time (the range of surveys spans from 1968 to 2021),
with 609 personal culture items capturing attitudes, beliefs, values,
self-assessments, self-descriptions, and behaviors (Alwin 2007). We
restricted the sample such that individuals between the ages of 18 and
79 are included without further elimination, and in all surveys, we used
all possible cases for which respondents provided responses. In the end,
the analyses that follow rely on a cross-national sample with a
cross-domain set of items to capture a broad range of individual
personal culture. Supplemental Materials A documents the list of all
variables used in the upcoming analyses.

@@

The estimation procedure begins with with Lersch's (2023) Life Course
Adaption Model, which formalizes survey responses at time \(t\) as a
function of individual-level random intercepts and slopes for survey
age. In some contexts, this model is called a mixed-effects growth curve
model. This model assumes a set of propositions about change that
reflect the theoretical debate to this point. First, consistent with the
settled disposition model, it assumes that people start the survey with
cultural differences, modeled as random intercepts for each respondent.
Second, it assumes that people change over time, taking the form of
random slopes for each respondent as a linear function of time. Third,
it assumes that people deviate around this baseline randomly over time,
reflecting ``fluctuation'' or short-term non-persistent change.
Formally, this can be written as

@@

As a first step in our analysis, we fit this LCAM to each of the 609
measures of personal culture outlined above. We then use these models to
derive two measures of the components we outlined above: pre-existing
interpersonal differences and intrapersonal change over time.

We estimate these using a form of variance decomposition. First, to
measure stable interpersonal differences, we calculate the sum of the
squared residuals derived from assuming that each person spends the
entire survey at the estimated midpoint of their LCAM
trajectory.\footnote{We use the midpoint estimate, rather than the value
  at the first wave, because this is the best measure of ``baseline''
  available under the assumption that the person does not change. The
  first wave measure alone contains an unknown amount of measurement
  error and transient fluctuations.} We then divide this value by the
total sum of squares and subtract this number from from 1, producing a
value we denote as \(V(D)\). This is the proportion of total variance
attributable to interpersonal differences. (This number is equivalent to
the conditional \(R^2\) value generated by a model with only random
intercepts for all respondents.)

Second, we calculate another sum of squared residuals generated by
allowing the predictions to vary by estimated LCAM trajectory. We divide
this number by the total sum of squares and subtract it from 1. We then
subtract \(V(D)\) from this value. This gives us a value we denote as
\(V(C)\), the additional proportion of variance accounted for by
intrapersonal change.

@@

The most notable result in the meta analysis is the negative coefficient
attached to the duration of years covered by the question. The longer a
question is observed, the less is the value of \(\omega\). Theories of
personal cultural change that link changes in personal culture to social
experiences, including the LCAM, suggest that the longer we observe
respondents, the more likely people are to undergo potentially
transformative experiences and therefore the more variance would be
attributable to intrapersonal change. Finding a negative coefficient
here seems to challenge that assumption.

@@

\leavevmode\vadjust pre{\hypertarget{ref-raftery1995}{}}%
Raftery, Adrian E. 1995. {``Bayesian {Model} {Selection} in {Social}
{Research}.''} \emph{Sociological Methodology} 25:111--63. doi:
\href{https://doi.org/10.2307/271063}{10.2307/271063}.
